TacticToe: Learning the Game of Tactics

### 2018-02-15 tactictoe_jv.tex
example commentary

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search~~\cite{montecarlo} (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kind of objects: tactics, theorems and list of goals.
\item The orthogonalization process that eliminates redundant tactics is
formally described.
\item A tactic abstraction method is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\todoi{More on evaluation??}
\item Evaluation of \tactictoe success rate relative to
the length of the original proofs.
\item User interaction is enhanced by the minimization and prettification
applied on the generated proof.
\end{itemize}

@@

In Example~\ref{ex:cs1},

@@

TacicToe is resourceful. Changing the shape of the feature of the theorems
to get better predictions.

\begin{example} \texttt{SURJ\_INJ\_INV} in theory \texttt{pred\_set}
\begin{lstlisting}[language=SMLSmall]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}
Human (2 milliseconds)
\begin{lstlisting}[language=SMLSmall]
REWRITE_TAC [IMAGE_SURJ] THEN
DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC THENL
  [irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
     REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
     REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
     FULL_SIMP_TAC std_ss [THE_DEF],
   ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
     RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
       [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC []]
\end{lstlisting}
\tactictoe (50 milliseconds)
\begin{lstlisting}[language=SMLSmall]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}
\end{example}

\begin{example} \texttt{LUPDATE\_APPEND2} in theory \texttt{rich\_list}
\begin{lstlisting}[language=SMLSmall]
!l1 l2 n x. LENGTH l1 <= n ==>
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}
Human (63 milliseconds)
\begin{lstlisting}[language=SMLSmall]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}
\tactictoe (17 milliseconds)
\begin{lstlisting}[language=SMLSmall]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}
\end{example}

fs is a \sml binding for FULL\_SIMP\_TAC (srw\_ss ()) and rw is short for ??.
If a discovered proof is short, fast, maintainable and
understandable, it could replace its original in the \holfour standard library.

To conclude, we also demonstrate the capabilities of \tactictoe on a goal
that did not have a proof in \holfour.

\todoi{commented out because the theorem exists in hoL4}.
%The first goal, in Example~, states that the set of 
%natural numbers is countable. \tactictoe starts from the definition of 
%countability and injection then instantiate the result with the trivial 
%injection and finalize the proof using the \metis. Here,
%\tactictoe is able to find an instantiation for an existential quantifier 
%because it is a common one. If a more complicated term was needed, \tactictoe 
%would try to find an alternative proof that do not rely on an original
%instantiation. This weakness could be mitigated by the introduction of term 
%prediction in a future version of \tactictoe.

%\begin{example}
%\begin{lstlisting}[language=SMLSmall]
%countable (UNIV:num set)
%let fun RWTAC thms = SRW_TAC [] thms in
%  RWTAC [countable_def, INJ_DEF] THEN Q.EXISTS_TAC `\\x.x` THEN METIS_TAC []
%end
%\end{lstlisting}
%\end{example}

In Example~, \todoi{TG to RK: I am not sure what the count 
constant is doing.}

\begin{example}
\begin{lstlisting}[language=SMLSmall]
count (n+m) DIFF count n = IMAGE ($+n) (count m)
SRW_TAC [ARITH_ss] [fetch "pred_set" "EXTENSION", EQ_IMP_THM] THEN
Q.EXISTS_TAC `x - n` THEN
SRW_TAC [ARITH_ss] []
\end{lstlisting}
\end{example}

>>>>

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search~~\cite{montecarlo} (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kind of objects: tactics, theorems and list of goals.
\item The orthogonalization process that eliminates redundant tactics is
formally described.
\item A tactic abstraction method is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe success rate relative to
the length of the original proofs.
\item User interaction is enhanced by the minimization and prettification
applied on the generated proof.
\end{itemize}

@@

In Example~\ref{ex:cs1}, the uniqueness of the greatest common divisor is 
proven. The human developer recognizes that it follows easily from two theorems.
Our prover \tactictoe is not able to find that \texttt{DIVIDES\_ANTISYM} is 
necessary directly from the shape of the original theorems so it rewrites the 
goal and split the goal into multiple cases until goal obligations are similar 
to \texttt{DIVIDES\_ANTISYM} to be predicted. As expected, \tactictoe's proof 
also takes much longer to check.

@@

In Example~\ref{ex:cs2}, we try to prove that for any surjective function $f$ 
from $s$ to $t$, there exists an injective function $g$ from $t$ to $s$ such 
that $f o g$ is the identity function. The human proof is quite complicated. In 
contrast, \tactictoe finds a proof that first expand the definition of 
injectivity and surjectivity and then call \metis. However, the simpler script 
does take much longer to execute due to the proof search happening inside 
\metis.

\begin{example}\label{ex:cs2} \texttt{SURJ\_INJ\_INV} in theory 
\texttt{pred\_set}
\begin{lstlisting}[language=SMLSmall]
!f s t. SURJ f s t ==> ?g. INJ g t s wedge !y. y IN t ==> (f (g y) = y)
\end{lstlisting}
Human (2 milliseconds)
\begin{lstlisting}[language=SMLSmall]
REWRITE_TAC [IMAGE_SURJ] THEN
DISCH_TAC THEN Q.EXISTS_TAC `THE o LINV_OPT f s` THEN
BasicProvers.VAR_EQ_TAC THEN REPEAT STRIP_TAC THENL
  [irule INJ_COMPOSE THEN Q.EXISTS_TAC `IMAGE SOME s` THEN
     REWRITE_TAC [INJ_LINV_OPT_IMAGE] THEN REWRITE_TAC [INJ_DEF, IN_IMAGE] THEN
     REPEAT STRIP_TAC THEN REPEAT BasicProvers.VAR_EQ_TAC THEN
     FULL_SIMP_TAC std_ss [THE_DEF],
   ASM_REWRITE_TAC [LINV_OPT_def, o_THM, THE_DEF] THEN
     RULE_ASSUM_TAC (Ho_Rewrite.REWRITE_RULE
       [IN_IMAGE', GSYM SELECT_THM, BETA_THM]) THEN ASM_REWRITE_TAC []]
\end{lstlisting}
\tactictoe (50 milliseconds)
\begin{lstlisting}[language=SMLSmall]
SRW_TAC [] [SURJ_DEF, INJ_DEF] THEN METIS_TAC []
\end{lstlisting}
\end{example}

In Example~\ref{ex:cs3}, we prove a theorem about lists which is domain where 
\tactictoe excels compared to ATPs. 
Given two list $l1$ and $l2$ where the length of $l1$ is less than a natural 
number $n$, the theorem states that performing an update on the $n^{th}$ 
element of the concatenation of $l1$ and $l2$ is the same as
performing an update on the $m^{th}$ element of $l2$ where $m = n - 
\mathit{length} l1$ followed by concatenation. Again, \tactictoe's proof is 
much more readable starting with applying induction on $l$. It solves the base 
case of the induction by rewriting and distinguish cases where $n=0$ or $n>0$ 
in the inductive hypothesis. It finalizes the proof using the definition of 
$LUPDATE$. Here, \tactictoe's proof is arguably smaller, faster and easier to 
understand and maintain. Such proofs after an expert review could replace the 
original human proofs in the \holfour repository.

\begin{example}\label{ex:cs3} \texttt{LUPDATE\_APPEND2} in theory 
\texttt{rich\_list}
\begin{lstlisting}[language=SMLSmall]
!l1 l2 n x. LENGTH l1 <= n ==>
  (LUPDATE x n (l1 ++ l2) = l1 ++ (LUPDATE x (n - LENGTH l1) l2))
\end{lstlisting}
Human (63 milliseconds)
\begin{lstlisting}[language=SMLSmall]
  rw[] THEN simp[LIST_EQ_REWRITE] THEN Q.X_GEN_TAC `z` THEN
  simp[EL_LUPDATE] THEN rw[] THEN simp[EL_APPEND2,EL_LUPDATE] THEN
  fs[] THEN Cases_on `z < LENGTH l1` THEN
  fs[] THEN simp[EL_APPEND1,EL_APPEND2,EL_LUPDATE]
\end{lstlisting}
\tactictoe (17 milliseconds)
\begin{lstlisting}[language=SMLSmall]
Induct_on `l1` THENL [SRW_TAC [] [],
  Cases_on `n` THENL [SRW_TAC [] [],
    FULL_SIMP_TAC (srw_ss ()) [] THEN METIS_TAC [LUPDATE_def]]]
\end{lstlisting}
\end{example}

In Example~\ref{ex:cs4}, we demonstrate the capabilities of 
\tactictoe on a goal that did not have a proof in \holfour.

\todoi{TG to RK: I am not sure what the count 
constant is doing. Feel free to add more examples.}

\begin{example}\label{ex:cs4}
\begin{lstlisting}[language=SMLSmall]
count (n+m) DIFF count n = IMAGE ($+n) (count m)
SRW_TAC [ARITH_ss] [fetch "pred_set" "EXTENSION", EQ_IMP_THM] THEN
Q.EXISTS_TAC `x - n` THEN
SRW_TAC [ARITH_ss] []
\end{lstlisting}
\end{example}
